Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-05-27 talk/dls2014/paper/paper.tex
text

We evaluate our system in a Python interpreter called
PyPy\footnote{www.pypy.org}. PyPy is an implementation of an
interpreter for the Python language. It has a special focus on speed,
as it provides a just-in-time (JIT) compiler to speed up applications
running on top of it. For comparison, we also do evaluation on other
Python interpreters:
\begin{description}
\item[CPython] is the reference implementation of the Python
  language. It is the most widely used interpreter for this language.
  The implementation uses a GIL for synchronisation in multi-threaded
  execution and it does not feature a JIT compiler.
\item[Jython] is an implementation of Python on top of the Java Virtual
  Machine (JVM). Instead of a GIL, this interpreter uses fine-grained
  locking for synchronisation. This enables true parallelism when
  executing code on multiple threads. In addition, its integration
  with the JVM provides it with a JIT compiler for faster execution.
\end{description}

@@

@@

There are several sources of extra memory requirements in our
STM system. First, we need to keep track of the state an object
is in. We do this using flags and an overflow number. Both currently
fit in an additional header of 4~bytes per object.

@@

Third, since we do copy on write, if all objects were to be
written to at the same time, all pages would need to be privatised
for all objects in the old object space. In that case we would
need the total amount of memory required by old objects multiplied
by $N+1$ (incl. the sharing segment).
\remi{maybe collect some statistics about pages privatised per segment}

\begin{itemize}
\item stm\_flags per object
\item read markers and other sections
\item private pages
\end{itemize}

maybe some memory usage graph over time

@@

\begin{itemize}
\item time taken by read \& write barriers
\item time spent committing \& aborting (maybe with different numbers
  of threads)
\item time in GC
\end{itemize}

>>>>

We evaluate our system in a Python interpreter called
PyPy\footnote{www.pypy.org} (version 2.3). PyPy is an implementation of an
interpreter for the Python language. It has a special focus on speed,
as it provides a just-in-time (JIT) compiler to speed up applications
running on top of it. For comparison, we compare between normal PyPy
using a GIL and a PyPy with STM, as well as between other Python interpreters:
\begin{description}
\item[CPython] (version 2.7.6) is the reference implementation of the Python
  language. It is the most widely used interpreter for this language.
  The implementation uses a GIL for synchronisation in multi-threaded
  execution and it does not feature a JIT compiler.
\item[Jython] (version 2.7b1) is an implementation of Python on top of
  the Java Virtual Machine (JVM). Instead of a GIL, this interpreter
  uses fine-grained locking for synchronisation. This enables true
  parallelism when executing code on multiple threads. In addition, its
  integration with the JVM provides it with a JIT compiler for faster
  execution.
\end{description}

@@

We performed all benchmarks on a machine with a Intel Core i7-4770
CPU~@3.40GHz (4 cores, 8 threads).  There are 16~GiB of memory
available and we ran them under Ubuntu 14.04 with a Linux 3.13.0
kernel. The STM system was compiled with a number of segments $N=4$
and a maximum amount of memory of 1.5~GiB (both are configurable at
compile time).

@@

There are several sources of extra memory requirements in our
STM system. First, we need to keep track of the state an object
is in. We do this using flags and an overflow number. Both currently
fit in a single additional header of 4~bytes per object.

@@

Third, since we do copy on write, if all objects were to be
written to at the same time, all pages would need to be privatised
for all objects in the old object space. In that case we would
need the total amount of memory required by old objects multiplied
by $N+1$ (incl. the sharing segment). Pages get re-shared during
major collections if possible.
\remi{maybe collect some statistics about pages privatised per segment}

\remi{The following discussion about richards mem usage does not
say that much... Also, RSS is not a good measure but it's hard to
get something better.}
In figure \ref{fig:richards_mem} we look at the memory usage of
one of our benchmarks called Richards\footnote{OS kernel simulation
benchmark}. The \emph{Resident Set Size} (RSS) shows the physical memory
assigned to the process. From it, we see that the process' memory
usage does not explode during the benchmark but actually stays pretty
much the same after start-up. Since it is the job of the OS to map
physical memory, this RSS number should be seen as a maximum. It is
possible that some of the memory is not required any more but still
assigned to our process.

The \emph{GC managed memory} counts all memory used in the old object
space including the memory required for private pages. The sharp drops
in memory usage come from major collections that free old objects and
re-share pages. Again the overall memory usage stays the same and
we see that in this benchmark we have around 1 major collection every
second.

For PyPy-STM the average memory requirement is 29~MiB and there are
$\sim 11$ major collections during the runtime. Normal PyPy with a GIL
grows its memory up to just 7~MiB and does not do a single major
collection in that time.

We are missing a memory optimisation to store small objects in a more
compact way, which is done by a normal PyPy not using STM.
Additionally, since normal PyPy uses a GIL, it does not need to
duplicate any data structures like e.g. the Nursery for each
thread. This, the missing optimisation, and the additional memory
requirements for STM explained above account for this difference.
\remi{I don't know how much sense it makes to go deeper. We will
improve this in the future, but right now this is the overall picture.}

@@

\remi{gs:segment prefix overhead is virtually none (maybe instruction cache)}
\remi{update numbers in pypy/TODO}

\begin{itemize}
\item time taken by read \& write barriers
\item time spent committing \& aborting (maybe with different numbers
  of threads; maybe split conflict detection and obj sync on commit)
\item time in GC
\end{itemize}
